---
title: "Getting Started with PhyloSubset"
author: "PhyloSubset"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PhyloSubset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(PhyloSubset)
library(ape)
set.seed(42)
```

## Introduction

PhyloSubset selects phylogenetically dispersed or clustered species subsets from a user-defined candidate pool and evaluates their distance-based dispersion and covariance-based dependence.

This vignette walks through the basic workflow.

## Step 1: Prepare a tree and candidate pool

```{r}
# Create a random tree for demonstration
tree <- rtree(30)
candidates <- tree$tip.label
```

## Step 2: One-step analysis with `phylo_subset()`

The main entry point runs the complete pipeline:

```{r}
res <- phylo_subset(
  tree = tree,
  candidates = candidates,
  size = 8,
  type = "dispersed",
  n_random = 200,
  cov_model = "BM",
  seed = 123
)
```

## Step 3: Examine results

```{r}
# Print summary
summary(res)

# View selected species
print(res$selected)

# View empirical p-values
print(res$empirical_p)
```

## Step 4: Visualize

```{r, fig.show = 'hold'}
# Plot tree with selected species
plot(res)

# Plot random baseline for MeanNND
plot(res, metric = "MeanNND")
```

## Step 5: Clustered subset

For comparison, select a clustered (high-dependence) subset:

```{r}
res_clust <- phylo_subset(
  tree = tree,
  candidates = candidates,
  size = 8,
  type = "clustered",
  n_random = 200,
  seed = 123
)

summary(res_clust)
```

## Step 6: Advanced step-by-step workflow

For more control, use the individual functions:

```{r}
# Compute distance matrix
D <- patristic_matrix(tree, candidates)

# Select dispersed subset
disp <- select_dispersed(
  dist_mat = D,
  candidates = candidates,
  size = 8,
  refine = TRUE,
  seed = 42
)

# Compute observed metrics
obs <- distance_metrics(D, disp$selected)

# Generate random baseline
rand <- random_baseline(
  dist_mat = D,
  candidates = candidates,
  size = 8,
  n = 200
)

# Compare to random
compare_to_random(obs, rand, type = "dispersed")

# Dependence diagnostics
V <- phylo_covariance(tree, candidates, model = "BM")
R <- cov_to_cor(V[disp$selected, disp$selected])
dependence_metrics(R)
```

## Covariance sensitivity analysis

Explore how dependence metrics change under different model assumptions:

```{r}
sens <- covariance_sensitivity(tree, disp$selected)
print(sens)
